---
title: "Mini Data-Analysis Deliverable 1"
output: github_document
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1.  Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

<!-- -->

    install.packages("devtools")
    devtools::install_github("UBC-MDS/datateachr")

2.  Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
```

3.  Make a repository in the <https://github.com/stat545ubc-2023> Organization. You can do this by following the steps found on canvas in the entry called [MDA: Create a repository](https://canvas.ubc.ca/courses/126199/pages/mda-create-a-repository). One completed, your repository should automatically be listed as part of the stat545ubc-2023 Organization.

# Instructions

## For Both Milestones

-   Each milestone has explicit tasks. Tasks that are more challenging will often be allocated more points.

-  Each milestone will be also graded for reproducibility, cleanliness, and coherence of the overall Github submission. 

- While the two milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as two chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

- The milestones will be equally weighted. 

## For Milestone 1

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-1.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work below --->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to the mini-analysis GitHub repository you made earlier, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 36 points: 30 for your analysis, and 6 for overall reproducibility, cleanliness, and coherence of the Github submission. 

# Learning Objectives

By the end of this milestone, you should:

-   Become familiar with your dataset of choosing
-   Select 4 questions that you would like to answer with your data
-   Generate a reproducible and clear report using R Markdown
-   Become familiar with manipulating and summarizing your data in tibbles using `dplyr`, with a research question in mind.

# Task 1: Choose your favorite dataset 

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

-   *apt_buildings*: Acquired courtesy of The City of Toronto's Open Data Portal. It currently has 3455 rows and 37 columns.

-   *building_permits*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 20680 rows and 14 columns.

-   *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

-   *flow_sample*: Acquired courtesy of The Government of Canada's Historical Hydrometric Database. It currently has 218 rows and 7 columns.

-   *parking_meters*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 10032 rows and 22 columns.

-   *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

-   *vancouver_trees*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

-   We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

-   Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 **(1 point)** Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

<!-------------------------- Start your work below ---------------------------->

1: cancer_sample
2: flow_sample
3: parking_meters
4: vancouver_trees

The reason why I chose to look further into these datasets is because these seemed to be the most interesting to work with, in my opinion. The cancer_sample data set allows me to look at the relationship between many different nuclei parameters and the diagnosis of breast cancer masses. The flow_sample dataset is interesting because it has a good mix of categorical and numeric variables that would make different types of analysis and plots possible to look at. The parking_meters dataset is interesting to me for similar reasons as the flow_sample dataset. It has information on the rates and time limits for parking meters in Vancouver. I thought it could be interesting to look at data pertaining to Vancouver and also to work with geographical data. Vancouver_trees data is also something I am considering because it too has a good mix of variable types. This dataset would also allow for interesting types of plots to be made to further investigate the relationship between variables.
<!----------------------------------------------------------------------------->

1.2 **(6 points)** One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*.

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you below, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

<!-------------------------- Start your work below ---------------------------->

```{r}
glimpse(cancer_sample)
glimpse(flow_sample)
glimpse(parking_meters)
glimpse(vancouver_trees)

```
I chose to use glimpse as my dplyr function because it shows me all of the columns and rows in the data frame, it tries to show as much data as possible. In this case I am able to see all of the variables and the different observations reported for that variable.
<!----------------------------------------------------------------------------->

1.3 **(1 point)** Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 1. What lead you to choose this one? Briefly explain your choice below.

<!-------------------------- Start your work below ---------------------------->
I choose cancer_sample because I think it will be interesting to work with a dataset looking at parameters between malignant and benign breast masses. I think the research questions formulated for this dataset will be interesting to investigate further. The dataset is also similar to the types of data I work with being in Neuroscience and will be a good learning tool for what I will be needing later on!
<!----------------------------------------------------------------------------->

1.4 **(2 points)** Time for a  final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with your dataset. Note it down below.

<!-------------------------- Start your work below ---------------------------->
Does the mean radius of nuclei differ in malignant versus benign breast masses?
<!----------------------------------------------------------------------------->

# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset 

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE under Task 3**.

2.1 **(12 points)** Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. 

Make sure that you're using dplyr and ggplot2 rather than base R for this task. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

1.  Plot the distribution of a numeric variable.
2.  Create a new variable based on other variables in your data (only if it makes sense)
3.  Investigate how many missing values there are per variable. Can you find a way to plot this?
4.  Explore the relationship between 2 variables in a plot.
5.  Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...
6.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
7.  Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
8.  Use a density plot to explore any of your variables (that are suitable for this type of plot).

2.2 **(4 points)** For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

<!-------------------------- Start your work below ---------------------------->
```{r}
ggplot(cancer_sample, aes(radius_worst)) +geom_histogram()+
ylab("Number of Observations")+
xlab("Largest Nuclei Size")

```

I chose to do **exercise 1** to give me a better idea of the size distribution of the "worst" or largest radius of nuclei present in the sample images. I can see what the largest nuclei size is across all observations as well as the most common size. Using a histogram allows me to easily see the distribution of sizes and which size tends to occur more frequently across all samples/observations. This **code** plots the radius_mean data in a histogram, I have also changed the labels for the graph to make the graph a bit easier to understand. I know there is an error message present but I think the histogram appears okay despite of the error.

```{r}
ggplot(cancer_sample, aes (diagnosis, radius_mean))+geom_jitter(alpha=0.5, width=0.1)+
  ylab("Mean Radius")+
  xlab("Diagnosis")
```

I chose to do **exercise 4** to explore the relationship between the mean radius of the nuclei from malignant or benign breast masses. This allows me to see individual data points and also the general radius size distribution within each diagnosis.This code plots the diagnosis and radius_mean variables on the same graph. Using a jitter geom allows for each data point to be seen. Overall, this **code** sorts the radius_mean based on the diagnosis and achieves what I was hoping to look at further. **Using alpha** in the code allows for some transparency in the data points and **width** allows for the points in the plot to not jitter too far away from each other.
```{r}
ggplot(cancer_sample, aes(diagnosis, radius_mean))+geom_boxplot(width=0.2)
```

```{r}
ggplot(cancer_sample, aes(diagnosis, texture_mean))+geom_boxplot(width=0.2)
```

In the earlier exercise, I used a jitterplot to see the distribution of the mean radius size of nuclei from differently diagnosed breast masses. I chose to do **exercise 6** next to be able to depict the average of radius means in each group. With this boxplot I am also able to see outliers present in each group which could be skewing the data. The outliers are more evident in the boxplot than they are in the jitterplot. I also included a boxplot looking at the mean texture of nuclei from malignant and benign masses to gain insight into that parameter as well. The **code** is also set to change the width of the bloxplots to a recommended setting of 0.2.
```{r}
ggplot(cancer_sample, aes(x=radius_mean))+
  geom_density (aes(fill=diagnosis, alpha=0.8))+
  scale_fill_manual(values=c("B"="cyan","M"="red"))
```

```{r}
ggplot(cancer_sample, aes(x=symmetry_mean))+
  geom_density (aes(fill=diagnosis, alpha=0.8))+
  scale_fill_manual(values=c("B"="cyan","M"="red"))
```

I chose to do exercise 8 next to plot the density of the mean radius and symmetry of nuclei for each diagnosis. This allows us to see how much overlap there is between the mean nuclei radius and diagnosis and also mean nuclei symmetry and diagnosis (second plot). Multiple densities in a single plot work well with a smaller number of categories (i.e. diagnosis). The symmetry of nuclei seems to overlap more than the radius for the different diagnosis. This plot also provides insight into the shape of the distribution. We are able to see 2 peaks for the mean radius of nuclei in malignant masses, this means there may be 2 frequently occurring values. The **code** allows me to plot the density curve of the radius_mean and using **fill** allows for the curves to be colored based on the diagnosis. **Alpha** is included in the code to make the curves slightly transparent so that the overlap can be seen more clearly. **The scale fill manual function** is to change the fill color so that benign is blue and malignant is red.
<!----------------------------------------------------------------------------->

# Task 3: Choose research questions 

**(4 points)** So far, you have chosen a dataset and gotten familiar with it through exploring the data. You have also brainstormed one research question that interested you (Task 1.4).  Now it's time to pick 4 research questions that you would like to explore in Milestone 2! Write the 4 questions and any additional comments below. 

<!--- *****START HERE***** --->
1. Does the mean radius size of nuclei present in the sample impact the diagnosis of the tissue (i.e. does it impact malignancy)?

I chose this question because I think it will be interesting to see whether the radius size of nuclei changes based on the diagnosis of the tissue type. It would be interesting to see if size is associated with malignancy.

2. Are any of the variables positively or negatively correlated with malignancy?

I chose this question because it will allow us to see which variable/parameters of the nuclei present in the sample image correlate with the type of diagnosis ultimately given to the breast mass.

3. Does the mean fractal dimension of the nuclei in the sample impact the malignancy diagnosis?

I chose this question because fractal dimension tells us information about accessible chromatin and gene activation and repression which can be linked to tumor progression. I think it will be interesting to investigate the relationship between these two variables in the dataset.

4. Can the mean fractal dimension predict the mean radius size of the nuclei present in the image, does an increase in fractal dimension result in an increase in radius size?

I chose this question to see whether there are any predictive factors that can be identified within this data set? I think it would be interesting to investigate whether these two factors are correlated and then look back at question 2 to see their impact/relationship with malignancy.

<!----------------------------->

# Overall reproducibility/Cleanliness/Coherence Checklist 

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors. An example of a major continuity error is having a data set listed for Task 3 that is not part of one of the data sets listed in Task 1. 

## Error-free code (3 points) 

For full marks, all code in the document should run without error. 1 point deduction if most code runs without error, and 2 points deduction if more than 50% of the code throws an error. 

## Main README (1 point)

There should be a file named `README.md` at the top level of your repository. Its contents should automatically appear when you visit the repository on GitHub.

Minimum contents of the README file:

-   In a sentence or two, explains what this repository is, so that future-you or someone else stumbling on your repository can be oriented to the repository.
-   In a sentence or two (or more??), briefly explains how to engage with the repository. You can assume the person reading knows the material from STAT 545A. Basically, if a visitor to your repository wants to explore your project, what should they know?

Once you get in the habit of making README files, and seeing more README files in other projects, you'll wonder how you ever got by without them! They are tremendously helpful.

## Output (1 point)

All output is readable, recent and relevant:

-   All Rmd files have been `knit`ted to their output md files. 
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

(0.5 point deduction if any of the above criteria are not met. 1 point deduction if most or all of the above criteria are not met.)

Our recommendation: right before submission, delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant. Then, after your final commit and push to Github, CHECK on Github to make sure that everything looks the way you intended!

## Tagged release (0.5 points)

You've tagged a release for Milestone 1. 

### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.