---
title: "Mini Data Analysis Milestone 2"
output: github_document
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome to the rest of your mini data analysis project!

In Milestone 1, you explored your data. and came up with research questions. This time, we will finish up our mini data analysis and obtain results for your data by: 

- Making summary tables and graphs 
- Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

We will also explore more in depth the concept of *tidy data.* 

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

# Instructions

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work here--->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to your mini-analysis GitHub repository, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 50 points: 45 for your analysis, and 5 for overall reproducibility, cleanliness, and coherence of the Github submission. 

**Research Questions**: In Milestone 1, you chose two research questions to focus on. Wherever realistic, your work in this milestone should relate to these research questions whenever we ask for justification behind your work. In the case that some tasks in this milestone don't align well with one of your research questions, feel free to discuss your results in the context of a different research question.

# Learning Objectives

By the end of this milestone, you should:

-   Understand what *tidy* data is, and how to create it using `tidyr`.
-   Generate a reproducible and clear report using R Markdown.
-   Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

# Setup

Begin by loading your data and the tidyverse package below:

```{r message=FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
```

# Task 1: Process and summarize your data 

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (1 point) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->
1. *Does the mean radius size of nuclei present in the sample impact the diagnosis of the tissue (i.e. does it impact malignancy)?*
2. *Are any of the variables positively or negatively correlated with malignancy?*
3. *Does the mean fractal dimension of the nuclei in the sample impact the malignancy diagnosis?*
4. *Can the mean fractal dimension predict the mean radius size of the nuclei present in the image, does an increase in fractal dimension result in an increase in radius size?*
<!----------------------------------------------------------------------------->

Here, we will investigate your data using various data manipulation and graphing functions.

### 1.2 (8 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

Also make sure that you're using dplyr and ggplot2 rather than base R. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Compute the proportion and counts in each category of one categorical variable across the groups of another categorical variable from your data. Do not use the function `table()`!

**Graphing:**

6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7. Make a graph where it makes sense to customize the alpha transparency.

Using variables and/or tables you made in one of the "Summarizing" tasks: 

8. Create a graph that has at least two geom layers. 
9. Create 3 histograms, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
Research question 1: Does the mean radius size of nuclei present in the sample impact the diagnosis of the tissue (i.e. does it impact malignancy)?

```{r}
observations <- cancer_sample %>% count(diagnosis) 
print(observations)

```

```{r}
ggplot(cancer_sample, aes(x=radius_mean))+
  geom_density (aes(fill=diagnosis, alpha=0.8))+
  scale_fill_manual(values=c("B"="cyan","M"="red"))
```
For research question 1, I chose **task 2** as it helps me get an idea of how many observations there are for each type of diagnosis. However, it does not directly contribute towards answering the research question.
**Task 7** allows me to see how much overlap there is between the mean nuclei radius and diagnosis. I see 2 peaks for the mean radius of nuclei in malignant masses, this means there may be 2 frequently occurring values. Alpha transparency in this case makes the curves slightly transparent so that the overlap can be seen more clearly.

Research question 2: Are any of the variables positively or negatively correlated with malignancy?

```{r}
radius_stats <- cancer_sample %>% 
  select(diagnosis, radius_mean) %>% 
  group_by(diagnosis) %>%
  summarise(.groups = "keep",
            min = min(radius_mean),
            mean = mean(radius_mean),
            median = median(radius_mean),
            max = max(radius_mean),
            range = max-min,
            stdev = sd(radius_mean)) #standard deviation
print(radius_stats)
```
```{r}
variable_stats <- cancer_sample %>%
group_by(diagnosis) %>%
summarise(across (where(is.numeric),
mean, na.rm=TRUE),
n=n())
print(variable_stats)
```
```{r}
ggplot(cancer_sample, aes(x=radius_mean, y=fractal_dimension_mean, colour=diagnosis)) +
  geom_point(alpha = 0.8)+
  geom_smooth(se=F, colour="black")
```
For research question 2, I chose to do **task 1** because it helps me gain a better understanding of different parameters and summary statistics of the nuclei radius mean size for each diagnosis. I chose to look at this variable because I think it may be correlated with the diagnosis received.  I also chose to summarise across all variables to see the means of each. This was helpful as it showed me which variables might be the most different between benign and malignant breast masses.
I chose to do **task 8** because this plot lets me explore and visualize the relationship between fractal dimension mean, radius size and diagnosis. Layering these visualizations gives me an initial look into whether these variables may be related in a linear fashion before running a linear regression.

Research question 3: Does the mean fractal dimension of the nuclei in the sample impact the malignancy diagnosis?

```{r}
nuclei_fractal_dimension_stats <- cancer_sample %>%
select(diagnosis, fractal_dimension_mean) %>% #picking radius mean vs. diagnosis
group_by(diagnosis) %>%
summarise(.groups = "keep",
min = min(fractal_dimension_mean),
Q1 = quantile(fractal_dimension_mean, 0.25), #1st quartile
mean = mean(fractal_dimension_mean),
median = median(fractal_dimension_mean),
Q3 = quantile(fractal_dimension_mean, 0.75), #3rd quartile
max = max(fractal_dimension_mean),
range = max-min,
stdev = sd(fractal_dimension_mean)) #standard deviation
print(nuclei_fractal_dimension_stats)

```
```{r}
ggplot(cancer_sample, aes (diagnosis, fractal_dimension_mean))+geom_jitter(alpha=0.5, width=0.1)+
  ylab("Nuclei Fractal Dimension Mean")+
  xlab("Diagnosis")+
  scale_y_log10()
```
For research question 3, I chose to do **task 1** to gain further insight into different summary statistics and parameters of a different variable I think might be correlated with diagnosis. This aids in answering the question of whether fractal dimension of nuclei impacts malignancy by allowing me to see  the summary statistics between diagnoses. I chose **task 6** for graphing because a jitter geom allows for each data point to be seen. **Using alpha** in the code allows for some transparency in the data points and **width** allows for the points in the plot to not jitter too far away from each other. I also used a **logarithmic scale** to plot the data, but since the values are very small it is less apparent.


Research question 4: Can the mean fractal dimension predict the mean radius size of the nuclei present in the image, does an increase in fractal dimension result in an increase in radius size?

```{r}
cancer_sample <- cancer_sample %>%
  mutate(radius_mean_category=case_when(
    radius_mean <= 8 ~ "Small",
    radius_mean <= 15 ~ "Medium",
TRUE~"Large"
  ))
print(cancer_sample)
```

```{r}
ggplot(cancer_sample, aes(x=fractal_dimension_mean))+
  geom_density (aes(fill=radius_mean_category, alpha=0.8))

```

For research question 4, I chose to do **task 3** to create a new variable where the numeric variable of radius mean is transformed into a categorical variable. I am able to classify the radius mean size as either small, medium or large. I chose graphing **task 7** to plot the density of the mean fractal dimension for each category of radius size. Multiple densities in a single plot work well with a smaller number of categories (i.e. radius mean category - small, medium or large). Here, I am able see whether larger fractal dimensions are present in nuclei with a larger mean radius. This plot also provides insight into the distribution of the data which would be otherwise harder to see in a table. **Fill** allows for the curves to be colored based on the radius category. **Alpha** is included in the code to make the curves slightly transparent so that the overlap can be seen more clearly.
<!----------------------------------------------------------------------------->

### 1.3 (2 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->
Based on the operations I ran above, I would say that I am much closer towards answering my research questions than I was prior to doing this. I was able to play around with the data and the variables to gain a deeper understanding of the relationships that are present. I do think that some of my research questions may have been too broad which makes it harder to properly summarise and graph the data to best fit the question. For broader research questions several plots may be required, and also more complicated analyses. The questions that are yielding interesting results are specific questions that I have asked. They are the ones that directly ask for the impact of on variable on another. This is in contrast to the question that asks if any of the variables are predictive of malignancy.

<!----------------------------------------------------------------------------->

# Task 2: Tidy your data 

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

-   Each row is an **observation**
-   Each column is a **variable**
-   Each cell is a **value**

### 2.1 (2 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have \>8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->
```{r}
head(cancer_sample)
```
Data seems tidy as each row is an observation, each cell is a value and each column is a variable. ID is an identifier variable as it is the patient ID. Therefore, each row is a new observation from a different patient. Let's investigate this further to double check.
```{r}
colnames(cancer_sample)
```
All of the columns are variables.

```{r}
glimpse(cancer_sample)
```
Each cell is a value. For diagnosis, the cells indicate either "M" or "B" for malignant or benign, respectively. This means that each cell is a value.

Based on the definition, the cancer_sample dataset is **tidy.**
<!----------------------------------------------------------------------------->

### 2.2 (4 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->

```{r}
head(cancer_sample)
```

```{r}
untidy_cancer_sample <- cancer_sample %>% 
   pivot_longer(cols = c(-ID, -diagnosis, -radius_mean_category),
                names_to = c("stat_type"), 
                values_to = "measurement") 
select(untidy_cancer_sample, stat_type, measurement, everything())
head(untidy_cancer_sample)
```
```{r}
tidy_cancer_sample<-untidy_cancer_sample%>%
  pivot_wider(names_from = stat_type, values_from = measurement) %>% 
  select(ID, diagnosis, radius_mean_category,everything())
head(tidy_cancer_sample)
```
<!----------------------------------------------------------------------------->

### 2.3 (4 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in the remaining tasks:

<!-------------------------- Start your work below ---------------------------->

1.  *Question 1: Does mean nuclei radius size differ depending on malignancy of the breast mass?*
2.  *New Question 2: Which single variable is the best predictor of malignancy is breast mass?*

<!----------------------------------------------------------------------------->

Explain your decision for choosing the above two research questions.

<!--------------------------- Start your work below --------------------------->
The reason why I chose these two questions is because they are specific and measurable. I am interested in seeing which single variable is important for diagnosis. These questions have yet to be answered but I believe I have gotten closer to the result through the plots and observations made with the data through the various exercises.
<!----------------------------------------------------------------------------->

Now, try to choose a version of your data that you think will be appropriate to answer these 2 questions. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelevant columns, etc.).

(If it makes more sense, then you can make/pick two versions of your data, one for each research question.) 

<!--------------------------- Start your work below --------------------------->

```{r}
question_2_data<- cancer_sample %>%
select(-c(ID)) %>%
mutate(malignancy = case_when(diagnosis == "M" ~ 1, TRUE ~ 0)) %>%
select(-diagnosis) %>%
group_by(malignancy) %>%
relocate (malignancy)
head(question_2_data)
```
Here, I have removed the ID column as it does not provide information that will be useful. I have assigned diagnosis with a numeric value to help with correlation analysis. "M" is assigned a 1 and "B" is given a 0. We, therefore, no longer need the diagnosis variable, so that has been removed too. The malignancy variable has also been moved to the front of the table so that it is easier to see.
```{r}
question_1_data<- question_2_data %>%
select(-c(contains("se"), contains("worst"), contains("category")))
head(question_1_data)
```
Here, I have removed columns containing "se", "category" or "worst" in order to be able to focus on the means of the variables. This will help answer question 1 but also help with investigating the relationship with other mean variables and malignancy.



# Task 3: Modelling

## 3.0 (no points)

Pick a research question from 1.2, and pick a variable of interest (we'll call it "Y") that's relevant to the research question. Indicate these.

<!-------------------------- Start your work below ---------------------------->

**Research Question**: Does mean nuclei radius size differ depending on malignancy of the breast mass?

**Variable of interest**: radius_mean

<!----------------------------------------------------------------------------->

## 3.1 (3 points)

Fit a model or run a hypothesis test that provides insight on this variable with respect to the research question. Store the model object as a variable, and print its output to screen. We'll omit having to justify your choice, because we don't expect you to know about model specifics in STAT 545.

-   **Note**: It's OK if you don't know how these models/tests work. Here are some examples of things you can do here, but the sky's the limit.

    -   You could fit a model that makes predictions on Y using another variable, by using the `lm()` function.
    -   You could test whether the mean of Y equals 0 using `t.test()`, or maybe the mean across two groups are different using `t.test()`, or maybe the mean across multiple groups are different using `anova()` (you may have to pivot your data for the latter two).
    -   You could use `lm()` to test for significance of regression coefficients.

<!-------------------------- Start your work below ---------------------------->
```{r}
data_test <- t.test(radius_mean ~ diagnosis, data=cancer_sample)
print(data_test)
```
Running a t-test on the means of the radius_mean size of nuclei within each diagnosis allows me to answer the quesiton of whether radius_mean is different between the two groups. This code chunk allows me to see the p-value, df, and means of each group.
<!----------------------------------------------------------------------------->

## 3.2 (3 points)

Produce something relevant from your fitted model: either predictions on Y, or a single value like a regression coefficient or a p-value.

-   Be sure to indicate in writing what you chose to produce.
-   Your code should either output a tibble (in which case you should indicate the column that contains the thing you're looking for), or the thing you're looking for itself.
-   Obtain your results using the `broom` package if possible. If your model is not compatible with the broom function you're needing, then you can obtain your results by some other means, but first indicate which broom function is not compatible.

<!-------------------------- Start your work below ---------------------------->
```{r}
library(broom)
tidy_data_test<- tidy(data_test)
print (data_test)
```
```{r}
#printing the p value
data_test$p.value
```
```{r}
#printing the mean
data_test$estimate
```

Here I was able to load the broom package and use its t-test function. I have also printed the p-value and mean outputs from the analysis. With this, I can see that there is a significant difference of radius mean size between malignant and bengin masses.
<!----------------------------------------------------------------------------->

# Task 4: Reading and writing data

Get set up for this exercise by making a folder called `output` in the top level of your project folder / repository. You'll be saving things there.

## 4.1 (3 points)

Take a summary table that you made from Task 1, and write it as a csv file in your `output` folder. Use the `here::here()` function.

-   **Robustness criteria**: You should be able to move your Mini Project repository / project folder to some other location on your computer, or move this very Rmd file to another location within your project repository / folder, and your code should still work.
-   **Reproducibility criteria**: You should be able to delete the csv file, and remake it simply by knitting this Rmd file.

<!-------------------------- Start your work below ---------------------------->
```{r}
output_directory <- "output"
file_name <- "milestone2_radius_summary.csv"
write_csv(radius_stats, here::here(output_directory, file_name))
```
I have written this summary table to a file called milestone2_radius_summary.
```{r}
test_summary <- read_csv(here::here(output_directory, file_name))
```
This allows me to check that the file has been written properly. I have read the fule and printed the contents.
<!----------------------------------------------------------------------------->

## 4.2 (3 points)

Write your model object from Task 3 to an R binary file (an RDS), and load it again. Be sure to save the binary file in your `output` folder. Use the functions `saveRDS()` and `readRDS()`.

-   The same robustness and reproducibility criteria as in 4.1 apply here.

<!-------------------------- Start your work below ---------------------------->
```{r}
saveRDS(data_test, here::here("output", "data_test.RDS"))
```

```{r}
readRDS(here::here("output", "data_test.RDS"))
```
After saving and reading the RDS files, we can see that it is the same model we made earlier. Therefore, the models have been properly written to the RDS files.
<!----------------------------------------------------------------------------->

# Overall Reproducibility/Cleanliness/Coherence Checklist 

Here are the criteria we're looking for.

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors. 

The README file should still satisfy the criteria from the last milestone, i.e. it has been updated to match the changes to the repository made in this milestone. 

## File and folder structure (1 points)

You should have at least three folders in the top level of your repository: one for each milestone, and one output folder. If there are any other folders, these are explained in the main README.

Each milestone document is contained in its respective folder, and nowhere else.

Every level-1 folder (that is, the ones stored in the top level, like "Milestone1" and "output") has a `README` file, explaining in a sentence or two what is in the folder, in plain language (it's enough to say something like "This folder contains the source for Milestone 1").

## Output (1 point)

All output is recent and relevant:

-   All Rmd files have been `knit`ted to their output md files. 
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

Our recommendation: delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant.

## Tagged release (0.5 point)

You've tagged a release for Milestone 2. 

### Attribution

Thanks to Victor Yuan for mostly putting this together.